Incorporating Periodic Variability in Hidden Markov Models for Animal Movement

### 2016-07-29 bmc_template/bmc_article.tex
need to center equations

%%% Load packages
%\usepackage{amsthm,amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

@@

@@

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transition probabilities -- in particular,
transition probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  
In addition to simulation-based examples,
we also re-analyze data from 
van de Kerk et al.\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
describe the movement and putative underlying
behavioural states of Florida panthers (\emph{Puma concolor coryi}).
% 
% %%\todo{we will clarify below that they actually used HSMMs} 
% %%to analyze hourly movement data
% %%from 18 Florida panthers (\emph{Puma concolor coryi}).
% 
van de Kerk et al.\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as six for individuals with a large amount
of available data); for reasons of computational 
practicality and biological interpretability, they restricted their
detailed analysis to models with only three underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\cite{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
states, but like \cite{kerk2015hidden} they chose to consider only 
models with three states. 
% (their criterion for choosing ...\todo{check \cite{zhao_knee_2008}!}).
As van de Kerk et al. \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
van de Kerk et al.'s model-based approach to
identifying the number of latent states.

@@

In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $ \mathbf Y = \obs_{1},..., \obs_{T}$ 
and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model
parameters \boldmath{$\theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:
\begin{eqnarray}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T} | \theta,\mathbf{X}_{1:T}) 
&=& P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \nonumber\\
&& \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\label{eq:HMMlik}
\end{eqnarray}

@@

\begin{eqnarray}
\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij} \nonumber
\end{eqnarray}

>>>>

%%% Load packages
%\usepackage{amsthm,amsmath}
\usepackage{amsmath}
%\RequirePackage{natbib}
%\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
%\RequirePackage{hyperref}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

@@

\

@@

\

Our primary goal in this paper is to introduce the use of
HMMs with temporally varying transition probabilities -- in particular,
transition probabilities that follow a diurnal cycle -- for modeling
animal movement recorded over long time scales.  
In addition to simulation-based examples,
we also re-analyze data from 
van de Kerk et al.\cite{kerk2015hidden}, who used temporally homogeneous hidden semi-Markov
models (HSMMs: an extension of HMMs that allow
flexible modelling of the distribution of \emph{dwell times}, the 
lengths of consecutive occupancy of a behavioural state) to
describe the movement and putative underlying
behavioural states of Florida panthers (\emph{Puma concolor coryi}).
% 
% %%\todo{we will clarify below that they actually used HSMMs} 
% %%to analyze hourly movement data
% %%from 18 Florida panthers (\emph{Puma concolor coryi}).
% 
\

van de Kerk et al.\cite{kerk2015hidden} found that the best-fitting HSMMs
incorporated a surprisingly large number of hidden
behavioural states (as many as six for individuals with a large amount
of available data); for reasons of computational 
practicality and biological interpretability, they restricted their
detailed analysis to models with only three underlying states.  In
contrast, most studies using HMM have chosen the
number of underlying states \emph{a priori}, typically using either two
\cite{schliehe-diecks_application_2012,mckellar_using_2014,langrock_flexible_2012,fryxell_multiple_2008}, or three states
\cite{dean2012behavioural,morales_extracting_2004,franke_prediction_2006}. 
In contrast, \cite{dean_behavioural_2013} evaluated models with up to 10
states, but like \cite{kerk2015hidden} they chose to consider only 
models with three states. 
% (their criterion for choosing ...\todo{check \cite{zhao_knee_2008}!}).
As van de Kerk et al. \cite{kerk2015hidden} comment, and as
we discuss further below, behavioural
repertoires with more than three distinct states are difficult to
interpret --- one possible reason that other authors have not adopted
van de Kerk et al.'s model-based approach to
identifying the number of latent states.

\

@@

In a HMM, the joint likelihood of \emph{emissions} (i.e., direct observations) $ \mathbf Y = \obs_{1},..., \obs_{T}$ 
and a hidden state sequence $\mathbf{Z}, \state_{t} \in \{1,...,n\}, t=1,...,T$, given model
parameters \boldmath{$\theta$} and covariates $\mathbf{X}_{1:T} =\cov_{1},..., \cov_{T}$, can be written as:

\begin{equation}
\begin{split}
P(\mathbf{Y}_{1:T},\mathbf{Z}_{1:T}|\boldsymbol \theta,\mathbf{X}_{1:T}) & =
P(\state_{1} \mid \cov_{1})P(\obs_{1} | \state_{1}, \cov_{1}) \cdot \\
&  \quad \prod\limits_{k=2}^{T}P(\state_{k} | \state_{k-1}, \cov_{k})P(\obs_{k} | \state_{k},\cov_{k})
\end{split}
\label{eq:HMMlik}
\end{equation}

@@

\

\begin{equation}
\begin{split}
\pi_{ij} & = \exp(\eta_{ij}(t))/\left(1+\sum_{j=2}^{n}\exp(\eta_{ij}(t))\right), \textrm{ for } j={2,...,n} \\
\pi_{i1} & = 1 - \sum_{j=2}^{n}\pi_{ij}
\end{split}
\end{equation}

We considered four different transition models for diurnal variation in behaviour, incorporating hour-of-day as a covariate following the general approach of Morales et al.\cite{morales_extracting_2004} of incorporating covariate dependence in the transition matrix.

\

\begin{description}
\item[Multiple block transition] Here we assume piecewise-constant transition probabilities. The transition probability $\pi_{ij}$ is a function of time (hour of day), where it is assigned to one of $M$ different time blocks:

$$
\eta_{ij}(t) = \sum_{m=1}^{M}a_{ijm} \delta_{m=t}
$$
where $a_{ijm}$ are parameters, and $\delta_{m=t}$ is a Kronecker delta ($\delta_{m=t}=1$ for the time block at the corresponding time $t$, and 0 otherwise).

\

\item[Quadratic transition model] We assume the elements of the linear predictor
are quadratic functions of hour. The quadratic model is not diurnally continuous, i.e. there is no constraint that forces $\eta_{ij}(0)=\eta_{ij}(24)$; imposing a diurnal continuity constraint would collapse the model to a constant.
$$
\eta_{ij}(t) = b_{ij1}+b_{ij2}\left(\frac{t}{24}\right)+b_{ij3}\left(\frac{t}{24}\right)^{2}
$$

\item[Sinusoidal transition model] A sinusoidal model with a period of 24 hours is identical in complexity to the quadratic model, but automatically satisfies the diurnal continuity constraint.
$$
\eta_{ij}(t)= b_{ij1}+b_{ij2}\cos\left(\frac{2\pi t}{24}\right)+b_{ij3}\sin\left(\frac{2\pi t}{24}\right)
$$

\item[Hourly model] Lastly, we extended the multi-block approach and assign  a different transition matrix for every hour of the day. This model is included for comparative purposes due to the large number of parameters in the model which makes it not really practical. We only fitted up to four states using the hourly model.  
\end{description}

\

Other periodic functions, such as Fourier series (the sinusoidal transition model augmented by additional sinusoidal components at higher frequencies) or periodic splines, could be useful directions for future exploration.

\

\subsection{Model evaluation}

\

We used the {\tt depmixS4} package
to fit covariate-dependent transition HMMs, 
simulate states and step lengths using the
estimated parameters, and estimate the most likely
states with the Viterbi
algorithm.

\

\newcommand{\dbic}{\ensuremath \Delta \textrm{BIC}}

We used three approaches to assess the fit of both time-homogeneous and time-inhomogeneous 
HMMs with 3 to 6 states to step-length data from the four of the thirteen Florida panthers 
with the most data ($> 9000 $ observations). (1) Comparing BICs to the optimal-BIC model 
within each type of transition complexity ($\dbic$ = BIC - min(BIC)) assesses the overall 
goodness of fit of each model type. (2) Comparing average step-length by hour of day for the 
observed data and for data simulated from the models shows how well a particular class of 
models can capture the diurnal variation in behaviour. (3) Comparing temporal autocorrelations
for the observed data and for data simulated from the models shows how well a particular class
of models captures serial correlation at both short and long scales.

\

Model complexity and the number of parameters increase as the number of latent states increase, FMM to HMM, and lastly, FMM and HMM incorporating temporal heterogeneity. The number of free parameters in an HMM can be generalized by summing up the number of free parameters of the three distinct components. Let $n$ be the number of hidden states and $k_{i}, k_{t}, k_{e}$ be the number of parameters describing the covariate-dependence of the prior distribution, transition function and emission distributions; that is, for a homogeneous model, $k=1$, while a single numeric covariate or a categorical predictor with two levels would give $k=2$. Then the number of free parameters of an HMM is:

\begin{equation}
\underbrace{k_{i}\cdot (n-1)}_{\text{Initial}} 
{+} \underbrace{k_{t}\cdot n\cdot (n-1)}_{\text{Transition}}
{+} \underbrace{ k_{e}\cdot n}_{\text{Emission}}
\label{eq:numparam}
\end{equation}
